---
title: "Statistical and Machine Learning SS26"
author: "Exercise Sheet 9"
output: pdf_document
output_dir: "/Users/Mara_1/Seafile/Statistical and Machine Learning/Exercises/Week 8 - Clustering/Prep"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Hierachical clustering

The `hclust()` function implements hierarchical clustering in `R`. \
In the following example we plot the hierarchical clustering dendrogram using three different rules for measuring the distance between *clusters* (not individual points) \
a) complete linkage clustering: merges groups only if all members are close \
b) single linkage clustering: merges groups if any pair of members is close \
c) average linkage clustering: averages over all cross-group pairs \ 

All of them are measured with Euclidean distance (most common choice - term for how we define 'distance' between points) as the dissimilarity measure.

```{r}
suppressWarnings(suppressMessages(library(tidyverse))) #load package, hide info messages
set.seed(2)
x <- matrix(rnorm(50 * 2), ncol =2) 
# generate 100 random numbers from a normal distribution, arrange them into a 50x2 matrix
# 50 rows/observations, 2 columns (coordinates)
# like 50 points of a 2 dimensional plane, all clustered around (0,0)
x[1:25, 1] <- x[1:25, 1] + 3 #select column 1 of rows 1-25 and add 3
# -> shifts first 25 points to the right by 3 units
# by doing that we create two seperate groups: 
# 25 points around x ~ 3, 25 points around x ~ 0
x[1:25, 2] <- x[1:25, 2] #y coordinates are left as they are
```

```{r}
plt_x <- as.data.frame(x) #converts matrix x into dataframe (ggplot2 requires df)
colnames(plt_x) <- c("V1", "V2") #name columns
plt_x$cluster = c(rep("C1",25), rep("C2", 25)) # adds new column `cluster` that 
# labels each point C1 or C2, depending on which group they belong to
ggplot(plt_x) + geom_point(aes(x = V1, y = V2, color = cluster))
# plots 50 points as scatter plot, we can clearly see that cluster C1 is scattered around 
# V1 ~ 3, while C2 is scattered around V1 ~ 0. Both have a V2 mean of 0 
# (we didn't shift any cluster in y direction)
```
Now everything we need has been set up and we can use the three different rules to measure the distance between clusters

a) Complete linkage clustering \
Complete linkage (farthest neighbor linkage) determines the distance between two clusters as the largest distance between any two points in the two clusters

1. Build 50×50 distance matrix (Euclidean distances)
   → 50 clusters, each 1 point

2. Find smallest value in matrix → merge those two points (e.g. A, B)
   → update matrix: replace A and B with cluster (A,B)
   → dist((A,B), C) = max(dist(A,C), dist(B,C))  (complete linkage rule)

3. Find smallest value in updated matrix → merge → update
   → dist((A,B), C) = max of all pairwise distances between cluster members

4. Repeat until 1 cluster remains

5. The full merge history = the dendrogram

https://www.statisticshowto.com/complete-linkage-clustering/

```{r}
hc.complete <- hclust(dist(x), method = "complete")
# dist()` function is used to compute the inter-observation Euclidean distance matrix
```

b) Single linkage clustering \
Single linkage (nearest neighbor linkage) determines the distance between two clusters as the shortest distance between any two points in the two clusters
```{r}
hc.single <- hclust(dist(x), method = "single")
```

c) Average linkage clustering \
Average linkage determines the distance between two clusters as the average distance between all pairs of points in the two clusters
```{r}
hc.average <- hclust(dist(x), method = "average")
```

Plot the dendrograms. The numbers at the bottom of the plot identify each observation.

```{r}
plot(hc.complete,main ="Complete Linkage", xlab = "", sub = "",cex = .9)
```
\
Complete Linkage: \
Two clearly separated, compact clusters \
The large gap between height 4 and 7.8 with no merges in between is exactly what we want to see: \
It confirms that cutting at 2 clusters is the right choice for this data \
Left cluster: points 1-25, right cluster: points 26-50 \
BUT It got 3 wrong! 44, 46, 16

```{r}
plot(hc.single,main ="Single Linkage", xlab = "", sub = "",cex = .9)
```
\
Single Linkage: \
Much smaller overall heights because it always uses the minimum distance \
Instead of two balanced groups forming independently, we see points being added one by one to a growing chain \

```{r}
plot(hc.average , main = "Average Linkage", xlab = "", sub = "",cex = .9)
```
\
Average linkage sits right between complete and single linkage \
It produces intermediate heights (makes sense since it uses the mean distance rather than max or min) \
Final merge happens later than with complete linkage (4 instead of 7.8), but it still produces two clear clusters like complete linkage \
Missclassifies 16, 44, 46 as well
For this dataset, complete linkage gives the clearest and most interpretable result

---

To determine the cluster labels for each observation associated with a given cut of the dendrogram, we can use the 'cutree()' function:
The second argument to 'cutree()' is the number of clusters we wish to obtain.


```{r}
h_complete <- cutree(hc.complete,2)
h_complete
# complete linkage cluster labels: generally seperates the observations into thir correct groups.
# algorithm missclassified only 3 out of 50 points (16, 44, 46)
# These three points are likely sitting close to the boundary between the two groups
```

```{r}
h_average <- cutree(hc.average , 2)
h_average
# average linkage missclassifies 16, 44, 46 as well - but correctly classifies the other 47 points!
```

```{r}
h_single <- cutree(hc.single,2)
h_single
# Single linkage identifies only one point as belonging to second cluster.
```


To scale the variables before performing hierarchical clustering of the observations, we use the 'scale()' function:

```{r}
xsc <- scale(x) #standardizes every column to mean = 0 and sd = 1
# Standardizing is crucial in real world data because variables with larger values or 
# larger variance would otherwise dominate the distance calculation and thus the clustering
hc.xsc <- hclust(dist(xsc), method = "complete")
plot(hc.xsc, main = "Hierarchical Clustering with Scaled Features")
# lower height because standardizing reduces the raw distances between points
# slightly messier dendrogram 
```
Here, the unscaled version was actually better for this specific dataset.
Unscaled data. x shift of +3, very clear separation of both groups, clearner dendrogram 
Scaled: shift gets compressed, separation reduced, messier


```{r}
h_complete_scaled <- cutree(hc.xsc,2)
h_complete_scaled
# takeaway of scaled vs unscaled: unscaled performed better! 
# for clustering with scaled features, there are more missclassifications!
```


## Evaluate the clustering results

In this example, we already know the true clusters. We can use `ARI` in package `mclust` to evaluate the clustering results.
ARI: Adjusted Rand Index, a metric that measures how well the clustering matches the true labels
ARI = 1 : perfect match, ARI = 0 : o better than random, ARI < 0 : worse than random

```{r}
suppressMessages(library(mclust))
```

```{r}
#code compares true lables (plt_x$cluster) with redicted cluster lables from
adjustedRandIndex(plt_x$cluster, h_complete) #complete linkage
adjustedRandIndex(plt_x$cluster, h_average) #average
adjustedRandIndex(plt_x$cluster, h_single) #single
adjustedRandIndex(plt_x$cluster, h_complete_scaled) #complete with scaled data
```
Which parameters showed best results?


# K-means clustering

In k means clustering, we must specify the number of clusters beforehand.
The function `kmeans()` performs K-means clustering in R. We begin with the same simple simulated example in previous section.

We now perform K-means clustering with K =2.

```{r}
km.out2 <- kmeans(x, 2, nstart = 20) 
#x: the data we generated, 2: number of clusters k
#nstart: runs algorithm 20 times with different random starting points, keeps the best result
km.out2
#Cluster sizes: 24, 26: Balanced, close to the true 25/25 split
# cluster means: centroids are exactly where we expect them: around 0;0 and 3;0
# 59% of the total variance in the data is explained by the cluster separation
# the remaining 41% is variance within the clusters themselves.
```

The cluster assignments of the 50 observations are contained in 'km.out$cluster'.

```{r}
km.out2$cluster #vector that contains the cluster assignments of the 50 obs.
```

```{r fig.width=12, fig.height=6}
#plot of data with each obs. colored according to its cluster assignment
par(mfrow = c(1, 2), mar = c(2, 2, 3, 1))
plot(x, col = (km.out2$cluster + 1), main = "K-Means Clustering Results with K = 2", 
     xlab = "",ylab ="",pch = 20, cex = 2)
# col = (km.out2$cluster + 1) makes points green and red, pch controls point shape, cex point size
plot(x, col = (as.numeric(as.factor(plt_x$cluster)) + 1), main = "True Cluster Assignments", 
     xlab = "", ylab = "", pch = 20, cex = 2)
# k means wrongly classifies 16,44,46 as well. but otherwise perfectly separated 
# the observations into two clusters, although we did not train the algorithm!
# same result as for complete linkage clustering!
```

How does this compare to hierarchical clustering?

```{r}
adjustedRandIndex(km.out2$cluster, h_complete) 
# how similar are k means clustering and complete linkage clustering's assignments to each other? 
#in this case, both methods produce the same cluster assignments, ARI=1 means they agree perfectly
```

```{r}
set.seed(1)
#we now run k-means with k=3 on our data, which only has two true clusters. 
# lets see how it performs:
km.out3 <- kmeans(x, 3) 
km.out3
```
K-means with k=3 on this data splits the data into three uneven groups: 10,21,19
Group 2 stayed intact (x and y around 0)
Group 1 (around x = 3) got split into an upper (new cluster 1) and lower (new cluster 3) group

WSS: Cluster 2 has the highest WSS — it's the largest and most spread out group

67% of variance is explained by the clusters (59% for k=2)
k=3 "explains" more, but only because it artificially splits a real group, creating fake separation
More clusters don't always mean better clustering if the extra clusters are artificial.
-> Choosing the right k is crucial! 

```{r}
plot(x, col = (km.out3$cluster + 1),
main = "K-Means Clustering Results with K = 3", xlab = "",ylab ="",pch = 20, cex = 2)
# which cluster was assigned which color?
```

## Choosing the number of clusters k

Silhouette index is a widely used method to evaluate clustering results. For each point $i \in C_I$, where $C_I$ is the cluster $I$, we first define

$$a(i) = \frac{1}{|C_I - 1|}\sum_{j \in C_I, i \neq j}d(i,j)$$

-> the mean distance between i and all other data points in the same cluster;

and 

$$b(i) = min_{J \neq I}\frac{1}{|C_J|} \sum{j \in C_J}d(i,j)$$
-> mean distance to members of the next closest cluster

The silhouette (value) of one data point $i$ is defined as 

$$s(i) = \frac{b(i) - a(i)}{max\{(b(i), a(i)\}} $$
numerator measures: how much better does this point fit in its own cluster compared to the next best cluster
denominator just picks whichever is larger between $a(i)$ and $b(i)$ and divides by it -> normalization!

We can also get the average silhouette width when using other algorithms:

With a hierarchical method (Complete linkage here):

```{r}
suppressWarnings(suppressMessages(library(cluster)))
dist.x <- dist(x) #computes and saves distance matrix
x.complete.link <- hclust(dist.x, method='complete') # complete linkage dendrogram

summary(silhouette(cutree(x.complete.link, k=2), dist(dist.x)))$avg.width
#cutree cuts dendrogram into 2 clusters and computes s(i) for every point, 
# then computes average silhouette width as one number

#same procedure for k = 3,4,5:
summary(silhouette(cutree(x.complete.link, k=3), dist(dist.x)))$avg.width
summary(silhouette(cutree(x.complete.link, k=4), dist(dist.x)))$avg.width
summary(silhouette(cutree(x.complete.link, k=5), dist(dist.x)))$avg.width
# Rule: pick k with the highest average silhouette width. Here: k = 2, silhouette width = 0.49 
```

With k-means:

```{r}
summary(silhouette(kmeans(x, centers=2, iter.max=100, nstart=25)$cluster, dist(x)))$avg.width
summary(silhouette(kmeans(x, centers=3, iter.max=100, nstart=25)$cluster, dist(x)))$avg.width
summary(silhouette(kmeans(x, centers=4, iter.max=100, nstart=25)$cluster, dist(x)))$avg.width
summary(silhouette(kmeans(x, centers=5, iter.max=100, nstart=25)$cluster, dist(x)))$avg.width
# same thing, k=2 with 0.483 highest average silhouette width
```

In each case, we choose the value of k associated with the LARGEST average silhouette width.


# Spectral clustering

Let’s look at how to implement spectral clustering and compare the results to a k-means approach.

```{r}
## Packages
library(tidyverse)
library(gridExtra)
library(ggplot2)
```

We first need to generate some data for clustering. We will generate three circular, uniform distributions and plot them:

```{r}
# Generate circular uniform distributions of data
set.seed(1748)

## Set min and max parameters for each distribution
min_vals <- c(0.05, 0.5, 1.5)
max_vals <- c(0.15, 0.8, 2.0)
x <- c() 
y <- c() 

## Generate distributions using parameters
for(i in 1:3){
  r <- sqrt(runif(250, min = min_vals[i], max = max_vals[i]))
  p <- runif(250)
  theta <- p * 2 * pi
  x_temp <- r * cos(theta)
  y_temp <- r * sin(theta)
  x <- c(x, x_temp)
  y <- c(y, y_temp)
}

## Create data frame
data <- as.data.frame(cbind(x, y))

## Plot coordinates
ggplot(data=data, aes(x=x, y=y)) + geom_point()

#result: 750 points total (250 per ring), arranged in 3 concentric rings
# clearly 3 clusters by eye
```

Our goal is to cluster the data. By sight, there are clearly three rings representing the three distributions of data. We will cluster the data using k-means first:

```{r}
# K-means Application
set.seed(1748)

## Run Kmeans algorithm with 3 clusters (sounds plausible as we have 3 rings)
km <- kmeans(data, 3) 

## Retrieve and store cluster assignments
kmeans_clusters <- km$cluster
kmeans_results <- cbind(data, cluster = as.factor(kmeans_clusters))
head(kmeans_results)
```

The output above represents the cluster each point is assigned to using k-means. We will now calculate the cluster assignments using spectral clustering:

```{r}
# Spectral Clustering

## Create Similarity matrix of Euclidean distance between points
S <- as.matrix(dist(data))

## Create Degree matrix
D <- matrix(0, nrow=nrow(data), ncol = nrow(data)) # empty nxn matrix

for (i in 1:nrow(data)) {
  
      # Find top 10 nearest neighbors using Euclidean distance
      index <- order(S[i,])[2:11]
      
      # Assign value to neighbors
      D[i,][index] <- 1 
}

# find mutual neighbors
D = D + t(D) 
D[ D == 2 ] = 1

# find degrees of vertices
degrees = colSums(D) 
n = nrow(D)


## Compute Laplacian matrix
# Since k > 2 clusters (3), we normalize the Laplacian matrix:
laplacian = ( diag(n) - diag(degrees^(-1/2)) %*% D %*% diag(degrees^(-1/2)) )


## Compute eigenvectors
eigenvectors = eigen(laplacian, symmetric = TRUE)
n = nrow(laplacian)
eigenvectors = eigenvectors$vectors[,(n - 2):(n - 1)]


set.seed(1748)
## Run Kmeans on eigenvectors
sc = kmeans(eigenvectors, 3)


## Pull clustering results
sc_results = cbind(data, cluster = as.factor(sc$cluster))
head(sc_results)
```

The output above represents the cluster each point is assigned to using spectral clustering. Let’s now graph the data and use colors to represent the cluster assignment of both k-means and spectral clustering:

```{r}
# Kmeans plot
kmeans_plot = ggplot(data = kmeans_results, aes(x=x, y=y, color = cluster)) + 
  geom_point() + 
  scale_color_manual(values = c('1' = "violetred2",
                                '2' ="darkorchid2",
                                '3' ="darkolivegreen2")) +
  ggtitle("K-Means Clustering") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        plot.title = element_text(hjust = 0.5), legend.position="bottom")

# Spectral Clustering plot
sc_plot = ggplot(data = sc_results, aes(x=x, y=y, color = cluster)) + 
  geom_point() + 
  scale_color_manual(values = c('1' = "violetred2",
                                '2' ="darkorchid2",
                                '3' ="darkolivegreen2")) +
  ggtitle("Spectral Clustering") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"),
        plot.title = element_text(hjust = 0.5), legend.position="bottom")

# Arrange plots
grid.arrange(kmeans_plot, sc_plot, nrow=1)
```

