---
title: "Statistical and Machine Learning SS26"
author: "Solution Exercise Sheet 5"
output: pdf_document
output_dir: "/Users/Mara_1/Seafile/Statistical and Machine Learning/Exercises/Week 5 - Kernel methods/Prep"

---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = getwd())
```


# 1)

a) Load and inspect the dataset "Melbourne House Prices".

```{r load}
melbdata <- read.csv("./Melbourne_housing_FULL.csv", header = TRUE) #load data
str(melbdata) #overview: nunber of obs&var, names of var
```

b) Plot the histogram of the distance from the city center as frequency and as density.

```{r} 
hist(na.omit(melbdata$Distance), main = "Distance to CBD", xlab = "Distance (km)") 
#na.omit removes missing values; default: freq=TRUE -> frequency; label plot; label x-axis

hist(na.omit(melbdata$Distance), breaks = 30,
     main = "Distance to CBD", xlab = "Distance (km)" )
# uses more bins -> histogram becomes more detailed 

hist(na.omit(melbdata$Distance), breaks = 30, freq = FALSE, 
     main = "Distance to CBD", xlab = "Distance (km)") 
#freq_FALSE -> density, area of all bars adds up to 1
```

c) Estimate the density of the Distance variable using four different kernel functions: Gaussian, Epanechnikov, Triangular, and Biweight. Use a fixed bandwidth of bw = 1 for all kernels. Plot the histogram of Distance as a density (breaks = 50). Overlay all four KDE curves on the histogram in different colors.

```{r}
d1 <- density(na.omit(melbdata$Distance), bw = 1, kernel = "gaussian") 
d2 <- density(na.omit(melbdata$Distance), bw = 1, kernel = "epanechnikov") 
d3 <- density(na.omit(melbdata$Distance), bw = 1, kernel = "triangular") 
d4 <- density(na.omit(melbdata$Distance), bw = 1, kernel = "biweight") 
#Each line estimates a smooth density curve from the same data, using a different kernel 
# shape but the same bandwidth bw = 1. 
# x axis: distance from city center; y axis:  the estimated density at each point

#Density at a point tells you how concentrated the data is there (not a probability!) it only 
# becomes an actual probability when computing the area under the curve over an interval.

#putting together histogram and 4 density curves
hist(na.omit(melbdata$Distance), freq = FALSE, breaks = 50, 
     main = "Effect of kernel function", xlab = "Distance to CBD (km)")
lines(d1$x, d1$y, col = "red") #takes x and y values of d1 (gaussian) and plots the curve
lines(d2$x, d2$y, col = "green2") 
lines(d3$x, d3$y, col = "purple") 
lines(d4$x, d4$y, col = "blue3") 
legend("topright",c("Gaussian","Epanechnikov", "Triangular", "Biweight"),
      lty = c(1, 1, 1, 1), col = c("red", "green2", "purple", "blue3"))
```

All four different kernel density curves look very similar!

d) Refit and plot the KDE using only the Gaussian kernel with three different bandwidths: bw = 0.5, bw = 1, and bw = 1.5.  

```{r} 
# only use gaussian kernel, differenet bandwidths 
d1 <- density(na.omit(melbdata$Distance), bw = 0.5, kernel = "gaussian") 
d2 <- density(na.omit(melbdata$Distance), bw = 1, kernel = "gaussian") 
d3 <- density(na.omit(melbdata$Distance), bw = 1.5, kernel = "gaussian") 

hist(na.omit(melbdata$Distance), freq = FALSE, breaks = 50, main = "Effect of bandwidth", 
     xlab = "Distance (km)")
lines(d1$x, d1$y, col = "red") 
lines(d2$x, d2$y, col = "green2") 
lines(d3$x, d3$y, col = "purple") 
legend("topright",c("bw=0.5", "bw=1", "bw=1.5"),
      lty=c(1,1,1,1), col=c("red", "green2", "purple"))
```

e) Use the density estimation to estimate the probability of finding a house for sale between 8 and 10km from the CBD

```{r} 
bw <- bw.bcv(na.omit(melbdata$Distance), lower = 0.01, upper = 10) 
# Instead of manually picking bw = 1 like before, bw.bcv() automatically finds 
# the optimal bandwidth using Biased Cross-Validation.
# it searches between 0.01 and 10 and picks the value that minimizes the estimation error.

d2 <- density(na.omit(melbdata$Distance), bw = bw, kernel = "gaussian") 
# computes KDE using optimal bandwidth we just calculated
f <- approxfun(d2) 
#d2 is just a table of values, approxfun turns these values into a function that 
# we can call at any x value in the range and gives the estimated density at x km

#each of the following methods compute the same thing: 
# the probability of finding a house for sale between 8 and 10km from the CBD
p <- integrate(f, 8, 10); p
#Directly integrates the function f between 8 and 10.

ind = which((d2$x>8)&(d2$x<10))
mean(d2$y[ind])*(10-8)
# Finds all grid points between 8 and 10, takes their average density, 
# multiplies by the interval width

mean(f(seq(8, 10, length.out = 100)))*(10-8)
#Creates 100 evenly spaced points between 8 and 10, evaluates f at each, 
# takes the average and multiplies by interval width. 
# (More precise than Method 2 because it uses more points.)


plot(d2, main = "Density estimation")
abline(v=8, col = "red")
abline(v=10, col = "red")
bw
```


# 2)

a) Load and inspect the data

```{r}
library(ISLR2) #load packages
library(caret)
data("Caravan")
attach(Caravan) #if we attach the data set, we don't have to write caravan$variable anymore

str(Caravan)
summary(Purchase)
348 / 5822 # only 6% purchased a caravan insurance!
```


```{r}
standardized.X <- scale(Caravan[, -86]) #transforms every variable to have mean 0; sd 1. 
# caravan data set, every row, every column BUT the 86th one (Purchase)

#by runnning the following two lines, we can see that the scales of two randomly 
# chosen variables are very different!
var(Caravan[, 1]) 
var(Caravan[, 2])

var(standardized.X[, 1]) #variance 1
mean(standardized.X[, 1]) #mean approx. zero
var(standardized.X[, 2])
# Every column of standardized.X has a standard deviation of one and a mean of zero.
```

Now we build the KNN model.

```{r}
# Split the observations into a test set and a training set
test <- 1:1000 # create a vector of row indices
train.X <- standardized.X[-test,] #takes all rows exept test rows (from standardized data)
test.X <- standardized.X[test,] #takes all test rows (1-1000)
train.Y <- Purchase[-test] #same thing for Y (standardized.X doesnt contain Purchase)
test.Y <- Purchase[test]

set.seed(123)

knn.fit <- knn3(x = train.X, train.Y, k = 1)
# Fits the KNN model on the training data
# x = train.X :  predictor matrix the model learns from
# train.Y : the true labels (Yes/No) for each training row
# k = 1 : use only the 1 nearest neighbor to make predictions
# -> for every new test point the model looks at the single closest 
# training point and assigns its label
# eg new customer -> Find the 1 closest person in training data 
# -> That person said "No" -> Predict "No"
# -> no real computation! It simply memorizes the entire training set

knn.pred <- predict(knn.fit, test.X, type = "class")
# predicts the class for each of the 1000 test observations 
# knn.fit: trained model
# test.X: 1000 test observations to predict
# type = class: returns predicted class label, rather than probabilities

# Evaluate its performance on the test data
mean(test.Y != knn.pred)  # computes the overall error rate of the KNN model on the test data
# true value/predicted value same or different? -> compute erreor rate 
# (number of wrong predictions/total predictions)
# Error rate on the test data is under 12 %

mean(test.Y != "No")
# only 6 % of customers purchased insurance, we could get the error rate down to 
# 6 % by always predicting No regardless of the values of the predictors!
# prediction error in naive model is lower than in KNN model 
```



```{r}
table(knn.pred , test.Y) # confusion matrix: rows predicted, columns actual values
10 / (68 + 10) 
# precision among predicted buyers: out of 78 predicted buyers, 10 will actually buy (12.8%)
```

```{r}
# Using K =3
knn.fit_K3 <- knn3(train.X, train.Y, k = 3) # same procedure but with k=3!
#For every test point, look at the 3 closest neighbors and take a majority vote
knn.predict_K3 <- predict(knn.fit_K3, test.X, type = "class")
table(knn.predict_K3 , test.Y)
#KNN is now more conservative -> predicts Yes much less often
5 / (5 + 19)
# but prediction quality improves: success rate increases to 21 %

# K = 5
knn.fit_K5 <- knn3(train.X, train.Y, k = 5) #k=5
knn.predict_K5 <- predict(knn.fit_K5, test.X, type = "class")
table(knn.predict_K5 , test.Y)
# as K increases, fewer people get flagged as buyers, higher precision among those flagged
4 / 15
# rate is 26.7 %

```

KNN: \
1. Standardize \
2. Overall accuracy is misleading for imbalanced data \
3. Confusion matrix & precision \
4. K tradeoff: Higher K = fewer but better predictions \



# 3)

Fit a Gaussian Process (GP) regression model to simulated data using the `gausspr()` function from the `kernlab` package
and interpret the posterior mean and uncertainty.

What's a GP? 
-> A distribution over functions. Instead of predicting a single value like linear regression, it predicts 
1) A mean: the best guess at each point
2) An uncertainty: how confident it is at each point
-> plot: line with confidence band around it 

Run the following code to generate the data.

```{r}
library(kernlab)
library(ggplot2)

set.seed(42)
f_true  <- function(x) sin(2 * pi * x) + 0.5 * sin(4 * pi * x) #true underlying function
n       <- 40
x_train <- sort(runif(n, 0, 1)) # 40 random x values between 0 and 1
y_train <- f_true(x_train) + rnorm(n, sd = 0.3) #true function values plus random noise
x_grid  <- seq(0, 1, length.out = 300) 
#grid of 300 points between 0 and 1 on which GP will make predictions on
```

a) Fit and plot the GP posterior mean

Fit a GP regression model using `gausspr()` with a squared-exponential (RBF) kernel (`kernel = "rbfdot"`), `sigma = 5`, and `var = 0.09`.
Predict on `x_grid` and plot the training data together with the posterior mean and the true function (dashed line).

```{r}
gp_fit <- gausspr(x_train, y_train, kernel = "rbfdot",
                  kpar = list(sigma = 5), var = 0.09)
# fits GP regression model. training data 
# (input(what we know)/output(what we want to predict))
# specify kernel function (gaussian; default choice)
# we look at kpar and var in c and d
# gp_fit then stores the fitted GP model

y_pred   <- predict(gp_fit, x_grid)
# uses the trained GP model to predict the posterior mean at every point in x_grid
# for every point in x_grid, the GP gives its best guess for y
# all values are stored in y

df_train <- data.frame(x = x_train, y = y_train) 
# combines x and y training data into data frame
df_grid  <- data.frame(x = x_grid, y_pred = as.vector(y_pred)) 
# data frame with x grid points and predicted values of y

ggplot() +
  geom_point(data = df_train, aes(x, y),
             colour = "steelblue", size = 2, alpha = 0.8) +
  geom_line(data = df_grid, aes(x, y_pred),
            colour = "firebrick", linewidth = 1) +
  geom_line(aes(x = x_grid, y = f_true(x_grid)),
            linetype = "dashed", colour = "grey40") +
  labs(title = "GP posterior mean (red) vs. true function (dashed)",
       x = "x", y = "y") +
  theme_minimal()

```



b) Obtain and plot posterior uncertainty

Refit the GP with `variance.model = TRUE`. Use `predict(..., type = "sdeviation")` to extract posterior standard
deviations and add $\pm 1.96$ credible bands to your plot.
Where is uncertainty largest, and why?

```{r}
gp_var <- gausspr(x_train, y_train, kernel = "rbfdot",
                  kpar = list(sigma = 5), var = 0.09,
                  variance.model = TRUE) 
#same code as in a) but it tells the GP to also track and return uncertainty 
# at each point, not just the mean

y_mean <- as.vector(predict(gp_var, x_grid)) 
#extracts posterior mean 

y_sd   <- as.vector(predict(gp_var, x_grid, type = "sdeviation")) 
# extracts posterior standard deviation at each grid point - how uncertain GP is

df_band <- data.frame( # computes data frame with:
  x  = x_grid, # x grid points
  mu = y_mean,  # their respective predicted mean
  lo = y_mean - 1.96 * y_sd, #  lower bound of band = mean - 1.96 × sd
  hi = y_mean + 1.96 * y_sd # upper bound of band = mean + 1.96 × sd
)

ggplot() +
  geom_ribbon(data = df_band, aes(x = x, ymin = lo, ymax = hi),
              fill = "firebrick", alpha = 0.2) +
  geom_line(data = df_band, aes(x = x, y = mu),
            colour = "firebrick", linewidth = 1) +
  geom_point(data = df_train, aes(x, y),
             colour = "steelblue", size = 2, alpha = 0.8) +
  labs(title = "GP posterior mean with 95% credible band",
       x = "x", y = "y") +
  theme_minimal()
```

The shaded band shows $\hat\mu(x) \pm 1.96\,\hat\sigma(x)$. \
---

c) Effect of the length-scale hyperparameter

Refit the GP with `sigma = 1` and `sigma = 50` (keeping `var = 0.09`).
Plot both posterior means together with the data and the true function.
Describe the effect of `sigma` on the fit.

```{r}
gp_s1  <- gausspr(x_train, y_train, kernel = "rbfdot",
                  kpar = list(sigma = 1),  var = 0.09)
gp_s50 <- gausspr(x_train, y_train, kernel = "rbfdot",
                  kpar = list(sigma = 50), var = 0.09)

df_comp <- data.frame(
  x       = x_grid,
  sigma1  = as.vector(predict(gp_s1,  x_grid)),
  sigma50 = as.vector(predict(gp_s50, x_grid))
)

ggplot() +
  geom_point(data = df_train, aes(x, y),
             colour = "steelblue", size = 2, alpha = 0.7) +
  geom_line(data = df_comp,
            aes(x, sigma1, colour = "sigma = 1"),  linewidth = 1) +
  geom_line(data = df_comp,
            aes(x, sigma50, colour = "sigma = 50"), linewidth = 1) +
  geom_line(aes(x = x_grid, y = f_true(x_grid)),
            linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("sigma = 1"  = "orange",
                                 "sigma = 50" = "purple")) +
  labs(title = "Effect of length-scale on GP fit",
       colour = "Kernel", x = "x", y = "y") +
  theme_minimal()
```

---

d) Effect of the noise variance

Using `sigma = 5`, refit the GP with `var = 0.001` and `var = 2`.
Plot the two posterior means. What does the noise variance parameter
control, and how does changing it affect the fit?

```{r}
gp_v001 <- gausspr(x_train, y_train, kernel = "rbfdot",
                   kpar = list(sigma = 5), var = 0.001)
gp_v2   <- gausspr(x_train, y_train, kernel = "rbfdot",
                   kpar = list(sigma = 5), var = 2)

df_noise <- data.frame(
  x    = x_grid,
  v001 = as.vector(predict(gp_v001, x_grid)),
  v2   = as.vector(predict(gp_v2,   x_grid))
)

ggplot() +
  geom_point(data = df_train, aes(x, y),
             colour = "steelblue", size = 2, alpha = 0.7) +
  geom_line(data = df_noise,
            aes(x, v001, colour = "var = 0.001"), linewidth = 1) +
  geom_line(data = df_noise,
            aes(x, v2,   colour = "var = 2"),     linewidth = 1) +
  scale_colour_manual(values = c("var = 0.001" = "darkgreen",
                                 "var = 2"     = "darkorange")) +
  labs(title = "Effect of noise variance on GP fit",
       colour = "Noise variance", x = "x", y = "y") +
  theme_minimal()
```






























