---
title: "Statistical and Machine Learning SS26"
author: "Exercise Sheet 6 (incl. solution)"
output: pdf_document
output_dir: "/Users/Mara_1/Seafile/Statistical and Machine Learning/Exercises/Week 6 - Tree based/Prep"

---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = getwd())
```

## Classification tree

Goal: Use classification trees to predict whether car seat sales are "High" or "Low"

The `Carseats` data set contains sales of child car seats at 400 different stores

`Sales` is a continuous variable (Unit sales (in thousands) at each location), we want to convert it into a binary variable

```{r}
library(tree) #used to construct classification and regression trees
library(ISLR2) #contains carseats data set
attach(Carseats)
str(Carseats)

High <- factor(ifelse(Sales <= 8, "No", "Yes")) 
# generate factor variable which takes on `yes` if the Sales variable exceeds 8, and `no` otherwise 

# merge factor variable `High` with the rest of the Carseats data
Carseats <-data.frame(Carseats , High)
```

We now use the `tree()` function to fit a classification tree in order to predict `High` using all variables but Sales. The syntax of the `tree()` function is quite similar to that of the `lm()` function.

```{r}
# fit model: dot to use all variables except sales as predictor 
tree.carseats <- tree(High ~.-Sales, Carseats) # high was derived from Sales, we 
#therefore exclude it because it doesn't contain new information 

# Check internal nodes in the tree, the number of terminal nodes, and the (training) error rate
summary(tree.carseats)
# shows us variables actually used (selected most informative 8): 
# Variables such as Education apparently do not contribute to predicting whether sales are High or Low
# the tree has 27 leaves: these are the endpoints where a final prediction ("Yes"/"No") is made. 
# training error rate is 9 % (36 out of 400 were misclassified). sounds good - 
# but we need to be skeptical as this model was tested on the same data it was trained on. 
# it has seen the answers already, so of course it knows what to do 
```

One of the most attractive properties of trees is that they can be graphically displayed. We use the `plot()` function to display the tree structure, and the `text()` function to display the node labels.

```{r fig.width=20, fig.height=10}
plot(tree.carseats)
text(tree.carseats , pretty =0) 
# pretty instructs R to include the category names for any qualitative predictors, 
# rather than simply displaying a letter for each category.
#left branch: yes, right branch: no
```

In order to properly evaluate the performance of a classification tree on these data, we must estimate the test error rather than simply computing the training error. We split the observations into a training set and a test set, build the tree using the training set, and evaluate its performance on the test data. The `predict()` function can be used for this purpose. In the case of a classification tree, the argument `type = "class"` instructs R to return the actual class prediction.

```{r}
set.seed(123)
train <- sample(1:nrow(Carseats), 200) 
Carseats.test <- Carseats[-train,]
High.test <- High[-train]
tree.carseats <- tree(High ~ .-Sales, Carseats, subset =train)
tree.pred <- predict(tree.carseats , Carseats.test , type = "class")
table(tree.pred , High.test) #confusion matrix

(87 + 65) / 200 
# correctly predicted no plus correctly predicted yes divided by total test obs (400 - 200)
# correct predictions for around 76 % of the locations in the test data set
```
Next, we consider whether pruning the tree might lead to improved results. The function `cv.tree()` performs cross-validation in order to determine the optimal level of tree complexity; cost complexity pruning `cv.tree()` is used in order to select a sequence of trees for consideration.

```{r}
set.seed(123)
cv.carseats <- cv.tree(tree.carseats , FUN = prune.misclass) 
# FUN =indicate that we want the classification error rate to guide the cross-validation
names(cv.carseats)
cv.carseats

#size: number of terminal nodes at each step (1: stump that predicts the same for all stores)
#dev: the cross-validation error at each size
#k: the complexity parameter (penalty for tree size, like lambda in regularization)
#method: just confirms "misclass" was used
```

We plot the error rate as a function of both size and k.
And we notice: the error barely changes between sizes 20 and 3. What does that suggest to you about whether you actually need a complex 20-node tree?
K vs. error plot: as we increase k (decrease size), the tree gets forced to be simpler and simpler, and eventually too simple $\rightarrow$ error rises


```{r}
par(mfrow = c(1, 2))
plot(cv.carseats$size , cv.carseats$dev,type ="b")
plot(cv.carseats$k, cv.carseats$dev,type ="b")
```

Prune the tree to obtain the nine-node tree.

```{r}
prune.carseats <- prune.misclass(tree.carseats , best = 9) 
#The error doesn't improve beyond size 9, so there's no benefit to keeping the extra nodes 
plot(prune.carseats)
text(prune.carseats , pretty =0)
tree.pred <- predict(prune.carseats , Carseats.test , type = "class")
table(tree.pred , High.test)

(86 + 65) / 200
# Now 77.5% of the test observations are correctly classified
# If we increase the value of `best`, 
# we obtain a larger pruned tree with lower classification accuracy
```

## Fit Regression Tree

We fit a regression tree to the `Boston` data set. First, we create a training set, and fit the tree to the training data

```{r fig.width=7, fig.height=5}
set.seed(123)
train <- sample(1:nrow(Boston), nrow(Boston) /2) 
#Creates a training set by randomly sampling half the observations
tree.boston <- tree(medv ~ ., Boston , subset =train) 
#fits regression (not classification) tree to predict meadian house value medv 
# using all other variables, but only on training split
summary(tree.boston) 
# instead of misclassification error rate we now see Residual mean deviance
# That's because this is a regression tree predicting a continuous value (medv), 
# not a category. We measure how far off our predictions are from the actual values

plot(tree.boston)
text(tree.boston , pretty =0)
# larger values of rm, or lower values of lstat, correspond to more expensive houses.

```

Now pruning the tree.

```{r}
set.seed(123)
cv.boston <- cv.tree(tree.boston) #Cross-validation to find the optimal tree size. 
# no FUN = prune.misclass because there's nothing to mislassify, 
# instead you measure how far off the predicted values are
plot(cv.boston$size , cv.boston$dev,type ="b")

prune.boston <- prune.tree(tree.boston , best = 5) # optimal tree size approx. 5. 
# again: prune.tree bc of regression, there are no observations that are misclassified
plot(prune.boston) # tree with 5 nodes instead of 8 - simpler and more interpretable
text(prune.boston , pretty =0)
```

```{r}
set.seed(123)
yhat <- predict(tree.boston , newdata = Boston[-train , ]) 
#Predicts medv for the test set sing the unpruned tree
boston.test <- Boston[-train,"medv"] #Extracts the actual medv values for the test set
plot(yhat , boston.test) #Plots predicted vs actual values
abline(0, 1) 
#draws a 45-degree line. A perfect model would have all points sitting exactly on that line. 
# more scatter -> worse predictions
mean((yhat - boston.test)^2) # 23.51 
sqrt(23.51) #4.85 -> in thousands: model's predictions are off by roughly $4,850 on average.
```

## Bagging and Random Forest

### Bagging

Here we apply bagging and random forests to the Boston data, using the `randomForest` package in R.

```{r}
library(randomForest)
set.seed(123)
bag.boston <- randomForest(medv ~ ., data = Boston, subset =train, mtry =12, importance = TRUE) 
# mtry = 12 indicates that all 12 predictors should be considered for each split 
# of the tree, not just a subset (as in random forest)
# importance = TRUE : Tells R to track which variables mattered most across all trees. 
# we can inspect this later with importance(bag.boston).

yhat.bag <- predict(bag.boston , newdata = Boston[-train , ]) #predicts on test set
plot(yhat.bag, boston.test) #plots predicted vs actual medv
abline(0, 1)
mean((yhat.bag - boston.test)^2) # and compute test MSE
# test set MSE associated with the bagged regression tree is 14.14, better than the single tree

bag.boston <- randomForest(medv ~ ., data = Boston , subset =train, mtry =12, ntree = 25) 
#same thing, but with 25 trees instead of default 500
yhat.bag <- predict(bag.boston , newdata = Boston[-train , ])
# test set MSE
mean((yhat.bag - boston.test)^2) # 14.63 -> higher MSE (Fewer trees = faster but noisier)
```

Growing a random forest proceeds in exactly the same way, except that we use a smaller value of the `mtry` argument.

```{r}
set.seed(123)
rf.boston <- randomForest(medv ~ ., data = Boston , subset = train , mtry = 6, importance = TRUE)
yhat.rf <- predict(rf.boston,newdata = Boston[-train,])
mean((yhat.rf -boston.test)^2)
# an improvement over bagging 
```

### Boosting

Here we use the `gbm()` function to fit boosted regression trees to the `Boston` data set

In boosting, each tree is grown from the mistakes of the previous tree. Specifically:

Tree 1 fits the data
Tree 2 focuses on the observations Tree 1 got wrong
Tree 3 focuses on what Tree 2 still got wrong
...and so on for 5000 trees

So the model learns gradually and corrects itself rather than building trees independently (as in bagging)

```{r}
library(gbm)
set.seed(123)
boost.boston <-gbm(medv ~ ., data = Boston[train , ], 
                   distribution = "gaussian", # since a regression problem
                   n.trees = 5000, 
                   interaction.depth = 4 # each individual tree is kept shallow (max 4 splits) 
                   )
summary(boost.boston)
# lstat and rm are the most important variables
```

```{r}
# Produce partial dependence plots for these two variables
plot(boost.boston , i = "rm") 
plot(boost.boston , i = "lstat")
```

These are partial dependence plots. They show the marginal effect of one variable on medv, after averaging out all other variables.

So for rm (avg rooms per house):

Houses with fewer than 5.5 rooms $\rightarrow$ predicted value flat around $20k
Between 5.5 and 7.5 rooms $\rightarrow$ steep increase
Above 7.5 rooms $\rightarrow$ flattens out around $35k

For lstat (% lower status population):

Low lstat $\rightarrow$ high house values ($30k)
As lstat increases, values drop steeply
Above  arouns 25% $\rightarrow$ flattens out around $16k

The interesting thing is the non-linear shape. A linear regression would only capture a straight line relationship. The boosted tree captures these plateaus and steep sections automatically.



We now use the boosted model to predict medv on the test set.

```{r}
set.seed(123)
yhat.boost <- predict(boost.boston, 
                      newdata = Boston[-train , ], n.tree = 5000)
mean((yhat.boost - boston.test)^2)
# better than the test MSE of bagging
```

Perform boosting with a different value of the shrinkage parameter $\lambda$.

```{r}
set.seed(123)
boost.boston <-gbm(medv ~ ., data = Boston[train , ], distribution = "gaussian", 
                   n.trees = 5000, interaction.depth = 4, shrinkage =0.02, verbose =F)
yhat.boost <- predict(boost.boston, newdata = Boston[-train , ], n.trees = 5000)
mean((yhat.boost - boston.test)^2)
```


















